### 2018-07-01 paper/5-implementations.tex
Work on S5.1

In the previous sections we discussed the types of build systems, and how they
can be broken down into two main components: a scheduler and a rebuilder.
In this section we make this abstract distinction concrete, by
implementing a number of build systems as a composition of a scheduler
and a rebuilder.  Our design is expressed by these types:
\begin{minted}[xleftmargin=10pt]{haskell}
type Scheduler c i ir k v = Rebuilder c ir k v -> Build c i k v
type Rebuilder c   ir k v = k -> v -> Task c k v -> Task (MonadState ir) k v
\end{minted}
So a \hs{Scheduler} is just a function that takes a \hs{Rebuilder} and uses
it to construct \hs{Build} system, by choosing which keys to rebuild in which order.
The \hs{Rebuilder} makes uses of the persistent build information \hs{ir}, while
the scheduler might augment that with further persistent information of its own,
yielding \hs{i}.

@@

The \hs{modTimeRebuilder} uses the pair
\hs{MakeInfo}~\hs{k}~\hs{=}~\hs{(@@now,}~\hs{modTimes)} as persistent
build information, carried by a \hs{MonadState} monad. This
\hs{MakeInfo} comprises the \emph{current time} \hs{now} and the map
\hs{modTimes}~\hs{::}~\hs{Map}~\hs{k}~\hs{Time} of \emph{file
modification times}. The rebuilder receives three arguments: a
\hs{key}, its current \hs{value}, and the \hs{task} of type
\hs{Task}~\hs{Applicative}~\hs{k}~\hs{v} that can be used to rebuild
the \hs{key} if necessary.  The rebuilder first decides if the
\hs{key} is \hs{dirty} by consulting \hs{modTimes}: if the key is not
found, that must mean it has never been built before; otherwise
\hs{modTimeRebuilder} can see if any of the \hs{task}'s dependencies
(computed by \hs{dependencies}) are out of date. If the key is
\hs{dirty}, we use \hs{run}~\hs{task} to rebuild it, and update the state with
the new modification time of the key\footnote{The real \Make
relies on the file system to track modification times, but we prefer
to make it explicit here.}; otherwise we can just return the current value.
We assume that the external system, which invokes the build system, updates
\hs{MakeInfo}~\hs{k} reflecting any file changes between successive builds.

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Make build system; stores current time and file modification times
type Time       = Integer
type MakeInfo k = (Time, Map k Time)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
make :: Ord k => Build Applicative (MakeInfo k) k v
make = topological modTimeRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on file modification times
modTimeRebuilder :: Ord k => Rebuilder Applicative (MakeInfo k) k v
modTimeRebuilder key value task = Task $ \fetch -> do
    (now, modTimes) <- get
    let dirty = case Map.lookup key modTimes of
            Nothing -> True
            time -> any (\d -> Map.lookup d modTimes > time) (dependencies task)
    if not dirty then return value else do
        put (now + 1, Map.insert key now modTimes)
        run task fetch
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A topological task scheduler
topological :: Ord k => Scheduler Applicative i i k v
topological rebuilder tasks target = execState $ forM_ order $ \key -> case tasks key of
    Nothing   -> return ()
    Just task -> do
        store <- get
        let newTask :: Task (MonadState i) k v
            newTask = rebuilder key (getValue key store) task
            fetch :: k -> State i v
            fetch k = return (getValue k store)
        newValue <- liftStore (run newTask fetch)
        modify $ putValue key newValue
  where
    deps k = case tasks k of { Nothing -> [@@]; Just task -> dependencies task }
    order  = fromJust $ topSort (reachable deps target) -- Assumes acyclic dependencies
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard graph algorithms (implementation omitted)
reachable :: Ord k => (k -> [k]) -> k -> Graph k
topSort   :: Ord k => Graph k -> Maybe [k] -- Returns Nothing on a cyclic graph
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Expand the scope of visibility of a stateful computation (implementation omitted)
liftStore :: State i a -> State (Store i k v) a
\end{minted}
\vspace{-3mm}
\caption{An implementation of \Make using our framework.}\label{fig-make-implementation}
\vspace{-3mm}
\end{figure}

\Make's scheduler, \hs{topological}, processes keys in a linear \hs{order} based on a
topological sort of the statically known dependency graph
(see~\S\ref{sec-parallelism} for parallel \Make). Our definition in
Fig.~\ref{fig-make-implementation} is polymorphic with respect to the type of
build information \hs{i} and is therefore compatible with any applicative
\hs{rebuilder}. The scheduler calls the supplied \hs{rebuilder} on every
\hs{key} in the \hs{order}, and runs the obtained \hs{newTask} to compute the
\hs{newValue} and updated build information \hs{newInfo}, which are subsequently
written to the store. The computation runs in the
\hs{State}~\hs{(@@Store}~\hs{i}~\hs{k}~\hs{v)} monad, similar to the \hs{busy}
build system (\S\ref{sec-general-build}), however since \hs{newTask} has access
only to the \hs{i} part of the \hs{Store}~\hs{i}~\hs{k}~\hs{v}, we need to use
\hs{runState} to execute it. The \hs{newTask} finds values of the \hs{key}'s
dependencies via the \hs{fetch} callback, which is defined to directly read the
\hs{store}.

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Bazel build system; stores constructive traces
bazel :: (Ord k, Hashable v) => Build Monad (CT k v) k v
bazel = restarting2 ctRebuilder -- implementation of 'restarting2' is omitted (22 lines)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- A rebuilder based on constructive traces
ctRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (CT k v) k v
ctRebuilder key value task = Task $ \fetch -> do
    cachedValues <- constructCT key (fmap hash . fetch) =<< get
    if value `elem` cachedValues then return value else case cachedValues of
        (cachedValue:_) -> return cachedValue
        _ -> do (newValue, deps) <- trackM task fetch
                put =<< recordCT key newValue deps (fmap hash . fetch) =<< get
                return newValue
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Cloud Shake build system, implementation of 'suspending' is given in Fig. 9
cloudShake :: (Ord k, Hashable v) => Build Monad (CT k v) k v
cloudShake = suspending ctRebuilder
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- CloudBuild build system, implementation of 'topological' is given in Fig. 7
cloudBuild :: (Ord k, Hashable v) => Build Applicative (CT k v) k v
cloudBuild = topological (adaptRebuilder ctRebuilder)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Convert a monadic rebuilder to the corresponding applicative one
adaptRebuilder :: Rebuilder Monad i k v -> Rebuilder Applicative i k v
adaptRebuilder rebuilder key value task = rebuilder key value $ Task $ run task
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Buck build system, implementation of 'topological' is given in Fig. 7
buck :: (Ord k, Hashable v) => Build Applicative (DCT k v) k v
buck = topological (adaptRebuilder dctRebuilder)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Rebuilder based on deep constructive traces, analogous to ctRebuilder
dctRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (DCT k v) k v
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Nix build system, implementation of 'suspending' is given in Fig. 9
nix :: (Ord k, Hashable v) => Build Monad (DCT k v) k v
nix = suspending dctRebuilder
\end{minted}
\vspace{-2mm}
\caption{\Bazel, \Cloud \Shake, \CloudBuild, \Buck and \Nix implemented in
our framework; \hs{restarting2} differs from \hs{restarting} used by \Excel
(Fig.~\ref{fig-excel-implementation}) in that it does not rely on the calc
chain.}
\label{fig-cloud-implementations}
\vspace{-4mm}
\end{figure}

>>>>

In the previous sections we discussed the types of build systems, and how they
can be broken down into two main components: a scheduler and a rebuilder.
In this section we make this abstract distinction concrete, by
implementing a number of build systems as a composition of a scheduler
and a rebuilder.  Our design is expressed by these types:
\begin{minted}[xleftmargin=10pt]{haskell}
type Scheduler c i ir k v = Rebuilder c ir k v -> Build c i k v
type Rebuilder c   ir k v = k -> v -> Task c k v -> Task (MonadState ir) k v
\end{minted}
So a \hs{Scheduler} is just a function that takes a \hs{Rebuilder} and uses
it to construct a \hs{Build} system, by choosing which keys to rebuild in which
order. The \hs{Rebuilder} makes uses of the persistent build information
\hs{ir}, while the scheduler might augment that with further persistent
information of its own, yielding \hs{i}.

@@

The \hs{modTimeRebuilder} uses the pair
\hs{MakeInfo}~\hs{k}~\hs{=}~\hs{(@@now,}~\hs{modTimes)} as persistent
build information, carried by a \hs{MonadState} monad. This
\hs{MakeInfo} comprises the \emph{current time} \hs{now} and the map
\hs{modTimes}~\hs{::}~\hs{Map}~\hs{k}~\hs{Time} of \emph{file
modification times}. The rebuilder receives three arguments: a
\hs{key}, its current \hs{value}, and the \hs{task} of type
\hs{Task}~\hs{Applicative}~\hs{k}~\hs{v} that can be used to rebuild
the \hs{key} if necessary.  The rebuilder first decides if the
\hs{key} is \hs{dirty} by consulting \hs{modTimes}: if the \hs{key} is not
found, that must mean it has never been built before; otherwise
\hs{modTimeRebuilder} can see if any of the \hs{task}'s dependencies
(computed by \hs{dependencies}) are out of date. If the \hs{key} is
\hs{dirty}, we use \hs{run}~\hs{task} to rebuild it, and update the state with
the new modification time of the \hs{key}\footnote{The real \Make
relies on the file system to track modification times, but we prefer
to make it explicit here.}; otherwise we can just return the current \hs{value}.
We assume that the external system, which invokes the build system, updates
\hs{MakeInfo}~\hs{k} reflecting any file changes between successive builds.

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Make build system; stores current time and file modification times
type Time       = Integer
type MakeInfo k = (Time, Map k Time)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
make :: Ord k => Build Applicative (MakeInfo k) k v
make = topological modTimeRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on file modification times
modTimeRebuilder :: Ord k => Rebuilder Applicative (MakeInfo k) k v
modTimeRebuilder key value task = Task $ \fetch -> do
    (now, modTimes) <- get
    let dirty = case Map.lookup key modTimes of
            Nothing -> True
            time -> any (\d -> Map.lookup d modTimes > time) (dependencies task)
    if not dirty then return value else do
        put (now + 1, Map.insert key now modTimes)
        run task fetch
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A topological task scheduler
topological :: Ord k => Scheduler Applicative i i k v
topological rebuilder tasks target = execState $ forM_ order $ \key -> case tasks key of
    Nothing   -> return ()
    Just task -> do
        store <- get
        let newTask :: Task (MonadState i) k v
            newTask = rebuilder key (getValue key store) task
            fetch :: k -> State i v
            fetch k = return (getValue k store)
        newValue <- liftStore (run newTask fetch)
        modify $ putValue key newValue
  where
    deps k = case tasks k of { Nothing -> [@@]; Just task -> dependencies task }
    order  = fromJust $ topSort (reachable deps target) -- Assumes acyclic dependencies
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard graph algorithms (implementation omitted)
reachable :: Ord k => (k -> [k]) -> k -> Graph k
topSort   :: Ord k => Graph k -> Maybe [k] -- Returns Nothing on a cyclic graph
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Expand the scope of visibility of a stateful computation
liftStore :: State i a -> State (Store i k v) a
liftStore x = do (a, newInfo) <- gets (runState x . getInfo)
                 modify (putInfo newInfo)
                 return a
\end{minted}
\vspace{-3mm}
\caption{An implementation of \Make using our framework.}\label{fig-make-implementation}
\vspace{-3mm}
\end{figure}

\Make's scheduler, \hs{topological}, processes keys in a linear \hs{order} based on a
topological sort of the statically known dependency graph
(see~\S\ref{sec-parallelism} for parallel \Make). Our definition in
Fig.~\ref{fig-make-implementation} is polymorphic with respect to the type of
build information \hs{i} and is therefore compatible with any applicative
\hs{rebuilder}. The scheduler calls the supplied \hs{rebuilder} on every
\hs{key} in the \hs{order}, and runs the obtained \hs{newTask} to compute the
\hs{newValue}. Note that \hs{newTask} has access only to the \hs{i} part of the
\hs{Store}~\hs{i}~\hs{k}~\hs{v}, but the rest of the \hs{do} block runs in the
\hs{State}~\hs{(@@Store}~\hs{i}~\hs{k}~\hs{v)} monad; we use the (unremarkable)
helper function \hs{liftStore} to fix the mismatch. The \hs{newTask} finds
values of the \hs{key}'s dependencies via the \hs{fetch} callback, which is
defined to directly read the \hs{store}.

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Bazel build system; stores constructive traces
bazel :: (Ord k, Hashable v) => Build Monad (CT k v) k v
bazel = restarting2 ctRebuilder -- implementation of 'restarting2' is omitted (22 lines)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A rebuilder based on constructive traces
ctRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (CT k v) k v
ctRebuilder key value task = Task $ \fetch -> do
    cachedValues <- constructCT key (fmap hash . fetch) =<< get
    if value `elem` cachedValues then return value else case cachedValues of
        (cachedValue:_) -> return cachedValue
        _ -> do (newValue, deps) <- trackM task fetch
                put =<< recordCT key newValue deps (fmap hash . fetch) =<< get
                return newValue
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Cloud Shake build system, implementation of 'suspending' is given in Fig. 9
cloudShake :: (Ord k, Hashable v) => Build Monad (CT k v) k v
cloudShake = suspending ctRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- CloudBuild build system, implementation of 'topological' is given in Fig. 7
cloudBuild :: (Ord k, Hashable v) => Build Applicative (CT k v) k v
cloudBuild = topological (adaptRebuilder ctRebuilder)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Convert a monadic rebuilder to the corresponding applicative one
adaptRebuilder :: Rebuilder Monad i k v -> Rebuilder Applicative i k v
adaptRebuilder rebuilder key value task = rebuilder key value $ Task $ run task
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Buck build system, implementation of 'topological' is given in Fig. 7
buck :: (Ord k, Hashable v) => Build Applicative (DCT k v) k v
buck = topological (adaptRebuilder dctRebuilder)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Rebuilder based on deep constructive traces, analogous to ctRebuilder
dctRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (DCT k v) k v
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Nix build system, implementation of 'suspending' is given in Fig. 9
nix :: (Ord k, Hashable v) => Build Monad (DCT k v) k v
nix = suspending dctRebuilder
\end{minted}
\vspace{-2mm}
\caption{\Bazel, \Cloud \Shake, \CloudBuild, \Buck and \Nix implemented in
our framework; \hs{restarting2} differs from \hs{restarting} used by \Excel
(Fig.~\ref{fig-excel-implementation}) in that it does not rely on the calc
chain.}
\label{fig-cloud-implementations}
\vspace{-4mm}
\end{figure}
